Ethnicity sensitive author disambiguation using semi-supervised learning

### 2015-08-27 paper/paper.tex
further imp
incremental disambiguation
Hussein's changes

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/pub-to-signature}
\caption{An example signature $s$ for "Doe, John". A \textit{signature} is
defined as unique piece of information identifying an author on a publication,
along with any other metadata that can be derived from it, such as publication
title, co-authors or date of publication.}
\label{fig:signature}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/workflow}
\caption{Pipeline for author disambiguation: (a)
signatures are \textit{blocked} to reduce computational complexity, (b) a linkage
function is built with supervised learning, (c) independently within each block, signatures
are grouped using agglomerative clustering.}
\label{fig:workflow}
\end{figure}

@@

The simplest and most common strategy for blocking, referred to hereon in as \emph{SFI},
groups signatures together if they share the same surname(s) and the same first
given name initial (\eg, \emph{SFI}("Doe, John") $==$ "Doe, J").
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

@@

More specifically, using census data extracted from \citep{rugglesintegrated}, we
build a logistic regression classifier for mapping the $(1,5)$-TF-IDF
representation of an author name to one of the seven ethnic groups.
Given a pair of signatures $(s_1, s_2)$, the proposed ethnicity
features are each computed as the estimated probability of $s_1$ to belong to
the corresponding ethnic group, multiplied by the estimated probability of
$s_2$ to belong to the same group. In doing so, the expectation is for the
linkage function to become sensitive to the actual origin of the authors
depending on the values of these features.
Indirectly, let us also note that these features hold discriminative power since
if author names are predicted to belong to different ethnic groups, then they
are also likely to correspond to distinct people.

@@

As part of this work, we developed a stand-alone application for author
disambiguation, publicly available online\footnote{\url{https://github.com/glouppe/beard}} for free reuse
or study.
Our implementation builds upon the Python scientific stack, making
use of the Scikit-Learn library \citep{scikitlearn} for the supervised learning
of a linkage function and of SciPy \citep{scipy} for clustering.
All components of the disambiguation pipeline have been designed to follow the
Scikit-Learn API \citep{scikitlearnAPI}, making them easy to maintain,
understand and reuse.
Our implementation is made to be efficient, exploiting parallelization when available and
making it ready for production environments.

@@

\begin{table}
\caption{Average precision, recall and f-measure scores on test folds.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c|c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}} \\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$\\
  \hline
  \hline
Baseline & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
\hline
Blocking = SFI & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841 & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850} & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
No name normalization & 0.9887 & 0.9697 & 0.9791 & 0.9931 & 0.9658 & 0.9793 \\
Name normalization & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Classifier = GBRT & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846} & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711 & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786 & 0.9850 & 0.9707 & 0.9778 \\
Training pairs = Blocked, balanced & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Clustering = Average linkage & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671 & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785 & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut & 0.9024 & 0.9828 & 0.9409 & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814 & 0.9940 & 0.9727 & 0.9832 \\
Block cut & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
  \hline
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Maximum recall of blocking strategies, and their number of blocks on ${\cal S}^\prime$.}
\label{table:blocking}
\centering
\begin{tabular}{|l|cc|c|}
  \hline
  \textbf{Blocking} & $R_\text{B3}^*$ & $R_\text{pairwise}^*$ & \# blocks \\
  \hline
  \hline
    SFI & 0.9828 & 0.9776 & 12978 \\
    Double metaphone & 0.9907 & 0.9863 & 9753 \\
    NYSIIS & 0.9902 & 0.9861 & 10857 \\
    Soundex & 0.9906 & 0.9863 & 9403 \\
  \hline
\end{tabular}
\end{table}

@@

Clustering together all signatures from the same block is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires almost no computation (\ie, both learning a linkage function
and running agglomerative clustering can be skipped -- only the blocking
function is needed to group signatures). In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give very good results,
with a slight advantage for block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (\eg, because it corresponds to a young
researcher at the beginning of his career), this therefore suggests that using
a cut-off threshold learned globally from the known data would in general give
satisfactory results, only marginally worse than if claimed signatures had been
known.

@@

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/pub-to-signature}
\caption{An example signature $s$ for "Doe, John". A \textit{signature} is
defined as unique piece of information identifying an author on a publication,
along with any other metadata that can be derived from it, such as publication
title, co-authors or date of publication \glnote{Remove a) and b.}}
\label{fig:signature}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/workflow}
\caption{Pipeline for author disambiguation: (a)
signatures are \textit{blocked} to reduce computational complexity, (b) a linkage
function is built with supervised learning, (c) independently within each block, signatures
are grouped using hierarchical agglomerative clustering.}
\label{fig:workflow}
\end{figure}

@@

The simplest and most common strategy for blocking, referred to hereon in as \textit{Surname and First Initial (SFI)},
groups signatures together if they share the same surname(s) and the same first
given name initial (\eg, \emph{SFI}("Doe, John") $==$ "Doe, J").
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

@@

More specifically, using census data extracted from \citep{rugglesintegrated},
we build a support vector machine classifier (using a linear kernel and
one-versus-all scheme for multi-class classification) for mapping the
 $(1,5)$-TF-IDF representation of an author name to one of the seven ethnic groups. Given a
pair of signatures $(s_1, s_2)$, the proposed ethnicity features are each
computed as the estimated probability of $s_1$ to belong to the corresponding
ethnic group, multiplied by the estimated probability of $s_2$ to belong to the
same group. In doing so, the expectation is for the linkage function to become
sensitive to the actual origin of the authors depending on the values of these
features. Indirectly, let us also note that these features hold discriminative
power since if author names are predicted to belong to different ethnic groups,
then they are also likely to correspond to distinct people.

@@

As part of this work, we developed a stand-alone application for author
disambiguation, publicly available online\footnote{\url{https://github.com/glouppe/beard}} for free reuse
or study.
Our implementation builds upon the Python scientific stack, making
use of the Scikit-Learn library \citep{scikitlearn} for the supervised learning
of a linkage function and of SciPy \citep{scipy} for clustering.
All components of the disambiguation pipeline have been designed to follow the
Scikit-Learn API \citep{scikitlearnAPI}, making them easy to maintain,
understand and reuse.
Our implementation is made to be efficient, exploiting parallelization when available and
making it ready for production environments.
It is also designed be runnable
in an incremental fashion, by enabling disambiguation only on specified blocks if desired,
instead of having to run the disambiguation process on the whole signature set.

@@

\begin{table}
\caption{Average precision, recall and f-measure scores on test folds.}
\label{table:results}
\centering
\begin{tabular}{|l|c c c|c c c|}
  \hline
                       & \multicolumn{3}{|c|}{\textbf{B3}} & \multicolumn{3}{|c|}{\textbf{Pairwise}} \\
  \textbf{Description} & $P$ & $R$ & $F$ & $P$ & $R$ & $F$\\
  \hline
  \hline
Baseline & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
\hline
Blocking = SFI & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
Blocking = Double metaphone & 0.9856 & 0.9827 & 0.9841 & 0.9927 & 0.9817 & 0.9871 \\
Blocking = NYSIIS & 0.9875 & 0.9826 & \textbf{0.9850} & 0.9936 & 0.9814 & \textbf{0.9875} \\
Blocking = Soundex & 0.9886 & 0.9745 & 0.9815 & 0.9935 & 0.9725 & 0.9828 \\
\hline
No name normalization & 0.9887 & 0.9697 & 0.9791 & 0.9931 & 0.9658 & 0.9793 \\
Name normalization & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Classifier = GBRT & 0.9901 & 0.9760 & 0.9830 & 0.9948 & 0.9738 & 0.9842 \\
Classifier = Random Forests & 0.9909 & 0.9783 & \textbf{0.9846} & 0.9957 & 0.9752 & \textbf{0.9854} \\
Classifier = Linear Regression & 0.9749 & 0.9584 & 0.9666 & 0.9717 & 0.9569 & 0.9643 \\
\hline
Training pairs = Non-blocked, uniform & 0.9793 & 0.9630 & 0.9711 & 0.9756 & 0.9629 & 0.9692 \\
Training pairs = Blocked, uniform & 0.9854 & 0.9720 & 0.9786 & 0.9850 & 0.9707 & 0.9778 \\
Training pairs = Blocked, balanced & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Clustering = Average linkage & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
Clustering = Single linkage & 0.9741 & 0.9603 & 0.9671 & 0.9543 & 0.9626 & 0.9584 \\
Clustering = Complete linkage & 0.9862 & 0.9709 & 0.9785 & 0.9920 & 0.9688 & 0.9803 \\
\hline
No cut & 0.9024 & 0.9828 & 0.9409 & 0.8298 & 0.9776 & 0.8977 \\
Global cut & 0.9892 & 0.9737 & 0.9814 & 0.9940 & 0.9727 & 0.9832 \\
Block cut & 0.9901 & 0.9760 & \textbf{0.9830} & 0.9948 & 0.9738 & \textbf{0.9842} \\
\hline
Combined best settings  & 0.9910 & 0.9780 & \textbf{0.9845} & 0.9959 & 0.9750 & \textbf{0.9853} \\
  \hline
\end{tabular}
\end{table}

@@

\begin{table}
\caption{Maximum recall $R_\text{B3}^*$ and $R_\text{pairwise}^*$ of blocking strategies, and their number of blocks on ${\cal S}^\prime$.}
\label{table:blocking}
\centering
\begin{tabular}{|l|cc|c|}
  \hline
  \textbf{Blocking} & $R_\text{B3}^*$ & $R_\text{pairwise}^*$ & \# blocks \\
  \hline
  \hline
    SFI & 0.9828 & 0.9776 & 12978 \\
    Double metaphone & 0.9907 & 0.9863 & 9753 \\
    NYSIIS & 0.9902 & 0.9861 & 10857 \\
    Soundex & 0.9906 & 0.9863 & 9403 \\
  \hline
\end{tabular}
\end{table}

@@

Clustering together all signatures from the same block is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires almost no computation (\ie, both learning a linkage function
and running agglomerative clustering can be skipped -- only the blocking
function is needed to group signatures). In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give very good results,
with a slight advantage for block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (\eg, because it corresponds to a young
researcher at the beginning of his career), this therefore suggests that either using
a cut-off threshold learned globally from the known data or using SFI would in general give
satisfactory results, only marginally worse than if claimed signatures had been
known.

\textit{Combined best settings.} \glnote{todo}

@@

Further improvements include ... \glnote{TODO}
